Formalizing Stack Safety as a Security Property

### 2023-02-03 submission5/core.tex
fix some unresolved refs, a few notes

Indeed, the call stack is an ancient~\cite{one06phrack} and
perennial~\cite{mitre-cwe,DBLP:conf/raid/VeendCB12,
  DBLP:conf/sp/SzekeresPWS13,
  DBLP:conf/sp/HuSACSL16,msrc-bluehat,chromium-security}
target for low-level attacks, frequently involving control-flow hijacking via corrupting
the return address, and memory corruption more generally.

@@

In \cref{sec:machine} we formalize the machine model, 
its {\em security semantics}, and the stack safety properties built on these. 
\Cref{sec:enforcement} describes the micro-policies that we test,
\cref{sec:test} the testing framework itself, and
\cref{sec:relwork,sec:future} discuss related and future work.

@@

A security extends a machine
with additional context about the identity of current and pending
functions (which act as security principals) and about the registers and memory they require
to be secure. This added context is purely notional;
it does not affect the real machine. The security context
evolves dynamically through the execution of security-relevant operations,
which include events like calls, returns, and frame manipulations.
Our security properties are phrased in terms of this context, often as predicates
on future states, e.g. of the form ``when control returns to the current function...'',
or relations on traces of future execution (hyper-properties).

@@

Whatever the mechanism for labeling transitions, we get a base
machine transition \(\mach \xrightarrow{\bar{\psi}, \obs} \mach'\), where \(\obs\) is
an event (see below) and \(\bar{\psi}\) is a list of security-relevant operations.
%We then lift this into a transition between pairs of machine states and contexts,
%\((\mach,\context) \stepstounder{\overline{\psi},\obs} (\mach', \context')\).

The set of security-relevant operations (\(\Psi\)) covered in this paper is given in
\cref{tab:psi}. Those that we explore in detail in \cref{sec:example}
are shown in white: calls, returns, and the allocation and deallocation of local memory.
In \cref{sec:machine}), we cover additional operations that add parameters to the examples,
to support various forms of memory sharing \ifexceptions , exceptions, \fi
and tail-call elimination. These are shown in light gray.
We give a proof-of-concept formalization of a more sophisticated, heap-like sharing model
based on pointer provenance in \cref{app:ptr}, represented by the dark grey
operations, but we do not test this.

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l | l |}
    \hline
    \(\psi \in \Psi\) & Parameters \\
    \hline
    \example \(\mathbf{call}\) & target address, argument registers \\
    \testing & stack arguments (base, offset \& size) \\
    \example \(\mathbf{return}\) & \\
    \example \(\mathbf{alloc}\) & offset \& size \\
    \testing & public flag \\
    \example \(\mathbf{dealloc}\) & offset \& size \\
    \testing \(\mathbf{tailcall}\) & (same as for \(\mathbf{call}\)) \\
    \hline
    \multicolumn{2}{|c|}{{\it Pointer provenance operations}} \\
    \hline
    \theory \(\mathbf{promote}\) & register, offset \& size \\
    \theory \(\mathbf{propagate}\) & source register/address \\
    \theory & destination register/address \\
    \theory \(\mathbf{clear}\) & target register/address \\
    \hline
  \end{tabular}
\end{center}
  \caption{Security-relevant Operations and their Parameters}
  \label{tab:psi}
\end{table}

@@

To characterize confidentiality, we borrow the idea of \emph{variant} states
from the theory of non-interference. Non-interference is a way of expressing
the flow of information through the system. If we begin with a state \(\mach\)
and modify it to \(\nach = \mach[\component \mapsto v]\) for arbitrary \(v\),
then execute both \(\mach\) and \(\nach\), any differences in the outputs they
produce must be the result of accessing \(\component\).

@@

In this section we introduce our security property definitions by means
of small code examples, using a simplified set of security-relevent operations for
calls, returns, and private allocations.
\Cref{fig:main} gives C code and possible corresponding compiled 32-bit RISC-V code
for a function {\tt main}, which
takes an argument {\tt secret} and initializes a local variable {\tt sensitive} to contain
potentially sensitive data.
Then {\tt main} calls another function {\tt f},
and afterward performs a test on {\tt sensitive} to decide whether
to output {\tt secret}.  Since {\tt sensitive} is initialized to 0,
the test should fail, and {\tt main} should instead output the return value of {\tt f}.
Output is performed by writing to the special global {\tt out},
and we assume that such writes are the only observable events in the system.

@@

Our security semantics attach a security context to the machine state,
which consists of a view \(V\) and a stack \(\sigma\) of pending activations' views.
\Cref{fig:exec1} shows how the security context evolves over the first few
steps of the program.  (The formal details of the security semantics are described in
\cref{sec:machine}, and the context evolution rules are formalized in \cref{fig:basicops}.)
Execution begins at the start of {\tt main}, where the program counter (\(\PCname\)) is zero,
and with the stack pointer (\(\SP\)) at address 1000.
State transitions are numbered and labeled with a list of security operations, written
\(\downarrow \overline{\psi}\) between steps.

@@

To begin with, what if {\tt f} returns to an unexpected place (i.e. \(\PCname \neq 20\) or
\(\SP \neq 980\))? We consider this to violate \(\wbcf\). \(\wbcf\) is a relationship between
call steps and their corresponding return steps: just after the return, the program
counter should be at the next instruction following the call,
and the stack pointer should be the same as it was before the call.
Both of these are essential. In \cref{subfig:WBCF}, the attacker adds
16 to the return address and then returns, thus bypassing the {\tt if}-test in the code and outputting
{\tt secret}.
In \cref{subfig:WBCF2}, the attacker returns with \(\SP' = 988\) instead of the
correct \(\SP = 980\). In this scenario, given the layout of {\tt main}'s frame,
\begin{center}
\begin{tabular}{| l | l | l | l | l |}
  \multicolumn{1}{r}{\(\SP \downarrow\)} &
  \multicolumn{2}{r}{\(\SP' \downarrow\)} \\
  \hline
  {\tt res} & {\tt sens} & {\tt sec} & \({\mathtt ra_1}\) & \({\mathtt ra_2}\) \\
  \hline
\end{tabular}
\end{center}

@@

\section{Formalization}
\label{sec:machine}

@@

\section{Expanding the Machine}

@@

\section{Enforcement}

@@

\subsection{Test Generation}
To use QuickChick, we developed random test-case generators that produce
(1) an initial RISC-V machine state, including the program to be executed;
(2)
  an initial policy state, tagging instructions and the stack regions appropriately
  for the micro-policy (see \cref{subsec:policies})
(3)
  a map from instructions to lists of their operations in the security semantics.

@@

\section{Related Work}

>>>>

Indeed, the call stack is an ancient~\cite{phrack96:smashingthestack} and
perennial~\cite{mitre-cwe,DBLP:conf/raid/VeendCB12,
  DBLP:conf/sp/SzekeresPWS13,
  DBLP:conf/sp/HuSACSL16,msrc-bluehat,chromium-security}
target for low-level attacks, frequently involving control-flow hijacking via corrupting
the return address, and memory corruption more generally.

@@

In \cref{sec:formal} we formalize the machine model, 
its {\em security semantics}, and the stack safety properties built on these. 
\Cref{sec:enforcement} describes the micro-policies that we test,
\cref{sec:testing} the testing framework itself, and
\cref{sec:relwork,sec:future} discuss related and future work.

@@

A security semantics extends a machine
with additional context about the identity of current and pending
functions (which act as security principals) and about the registers and memory they require
to be secure. This added context is purely notional;
it does not affect the real machine. The security context
evolves dynamically through the execution of security-relevant operations,
which include events like calls, returns, and frame manipulations.
Our security properties are phrased in terms of this context, often as predicates
on future states, e.g. of the form ``when control returns to the current function...'',
or as relations on traces of future execution (hyper-properties).

@@

Whatever the mechanism for labeling transitions, we get a base
machine transition \(\mach \xrightarrow{\bar{\psi}, \obs} \mach'\), where \(\obs\) is
an event (see below) and \(\bar{\psi}\) is a list of security-relevant operations.
%We then lift this into a transition between pairs of machine states and contexts,
%\((\mach,\context) \stepstounder{\overline{\psi},\obs} (\mach', \context')\).
The set of security-relevant operations (\(\Psi\)) covered in this paper is given in
\cref{tab:psi}. Those that we explore in detail in \cref{sec:example,sec:formal}
are shown in white: calls, returns, and the allocation and deallocation of local memory.
In \cref{sec:extensions}, we cover additional operations and and parameters 
to support various forms of memory sharing \ifexceptions , exceptions, \fi
and tail-call elimination; these are shown in light gray. These are also
the operations we test in \cref{sec:testing}.
We give a proof-of-concept formalization of a more sophisticated, heap-like sharing model
based on pointer provenance in \cref{app:ptr}, represented by the dark gray
operations, but we do not test this.

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l | l |}
    \hline
    \(\psi \in \Psi\) & Parameters \\
    \hline
    \example \(\mathbf{call}\) & target address, argument registers \\
    \testing & stack arguments (base, offset \& size) \\
    \example \(\mathbf{return}\) & \\
    \example \(\mathbf{alloc}\) & offset \& size \\
    \testing & public flag \\
    \example \(\mathbf{dealloc}\) & offset \& size \\
    \testing \(\mathbf{tailcall}\) & (same as for \(\mathbf{call}\)) \\
    \hline
    \multicolumn{2}{|c|}{{\it Pointer provenance operations}} \\
    \hline
    \theory \(\mathbf{promote}\) & register, offset \& size \\
    \theory \(\mathbf{propagate}\) & source register/address \\
    \theory & destination register/address \\
    \theory \(\mathbf{clear}\) & target register/address \\
    \hline
  \end{tabular}
\end{center}
\caption{Security-relevant operations and their parameters.}
  \label{tab:psi}
\end{table}

@@

To characterize confidentiality\apt{and sometimes integrity}, we borrow the idea of \emph{variant} states
from the theory of non-interference. Non-interference is a way of expressing
the flow of information through the system. If we begin with a state \(\mach\)
and modify it to \(\nach = \mach[\component \mapsto v]\) for arbitrary \(v\),
then execute both \(\mach\) and \(\nach\), any differences in the outputs they
produce must be the result of accessing \(\component\).

@@

In this section we introduce our security property definitions by means
of small code examples, using a simplified set of security-relevant operations for
calls, returns, and private allocations.
\Cref{fig:main} gives C code and possible corresponding compiled 32-bit RISC-V code
for a function {\tt main}, which
takes an argument {\tt secret} and initializes a local variable {\tt sensitive} to contain
potentially sensitive data.
Then {\tt main} calls another function {\tt f},
and afterward performs a test on {\tt sensitive} to decide whether
to output {\tt secret}.  Since {\tt sensitive} is initialized to 0,
the test should fail, and {\tt main} should instead output the return value of {\tt f}.
Output is performed by writing to the special global {\tt out},
and we assume that such writes are the only observable events in the system.

@@

Our security semantics attach a security context to the machine state,
which consists of a view \(V\) and a stack \(\sigma\) of pending activations' views.
\Cref{fig:exec1} shows how the security context evolves over the first few
steps of the program.  (The formal details of the security semantics are described in
\cref{sec:formal}, and the context evolution rules are formalized in \cref{fig:basicops}.)
Execution begins at the start of {\tt main}, where the program counter (\(\PCname\)) is zero,
and with the stack pointer (\(\SP\)) at address 1000.
State transitions are numbered and labeled with a list of security operations, written
\(\downarrow \overline{\psi}\) between steps.

@@

To begin with, what if {\tt f} returns to an unexpected place (i.e. \(\PCname \neq 20\) or
\(\SP \neq 980\))? We consider this to violate \(\wbcf\). \(\wbcf\) is a relationship between
call steps and their corresponding return steps: just after the return, the program
counter should be at the next instruction following the call,
and the stack pointer should be the same as it was before the call.
Both of these are essential. In \cref{subfig:WBCF}, the attacker adds
16 to the return address and then returns, thus bypassing the {\tt if}-test in the code and outputting
{\tt secret}.
In \cref{subfig:WBCF2}, the attacker returns with \(\SP' = 988\) instead of the
correct \(\SP = 980\). In this scenario, given the layout of {\tt main}'s frame,
\begin{center}
\begin{tabular}{| l | l | l | l | l |}
  \multicolumn{1}{r}{\(\SP \downarrow\)} &
  \multicolumn{2}{r}{\(\SP' \downarrow\)} \\
  \hline
  {\tt res} & {\tt sens} & {\tt sec} & \(\mbox{\tt ra}_1\) & \(\mbox{\tt ra}_2\) \\
  \hline
\end{tabular}
\end{center}

@@

\section{Formalization}
\label{sec:formal}

@@

\section{Expanding the Machine}
\label{sec:extensions}

@@

\section{Enforcement}
\label{sec:enforcement}

@@

\subsection{Test Generation}
To use QuickChick, we developed random test-case generators that produce
(1) an initial RISC-V machine state, including the program to be executed;
(2)
  an initial policy state, tagging instructions and the stack regions appropriately
  for the micro-policy (see \cref{sec:enforcement})
(3)
  a map from instructions to lists of their operations in the security semantics.

@@

\section{Related Work}
\label{sec:relwork}
